---
title: "Distributional Regression Illustration"
author: "Tim Wolock"
date: "05/03/2021"
output:
  html_document:
    toc: true
    number_sections: false
    toc_float: true
    df_print: "tibble"
bibliography: references.bib
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, message = F, warning = F,
               fig.align = 'center', fig.height = 3)
```

In this document, which accompanies Wolock *et al.* (2021), we will demonstrate how to fit a distributional model with a sinh-arcsinh likelihood in `brms` [@bürkner2018]. First, we will outline how to implement a custom family in `brms`, then we will simulate sinh-arcsinh distributed data with moments that vary with data, and finally, we will fit one conventional regression model and two distributional models.

The likelihood we will be using comes from @jones2009. The density of a sinh-arcsinh random variable, $X$, is

$$
p(x\;|\;\mu, \sigma, \epsilon, \delta) = \frac{1}{\sigma\sqrt{2\pi}}\frac{\delta C_{\epsilon, \delta}(x_z)}{\sqrt{1+x_z^2}} \exp\left[-\frac{S_{\epsilon,\delta}(x_z)^2}{2}\right],
$$

where $x_z = (x-\mu)/\sigma$, $S_{\epsilon,\delta}(x) = \sinh(\epsilon + \delta\mathrm{asinh}(x))$, and $C_{\epsilon,\delta}=\cosh(\epsilon + \delta\mathrm{asinh}(x))$. Both $\sigma$ and $\delta$ must be greater than zero. This distribution is considerably more flexible than the normal distribution.

## Setup

To step through this analysis yourself, clone or download this repository, open `distreg-illustration.Rproj` in RStudio, and open the `.Rmd` file in the RStudio editor. Be sure to have the following package installed:

1.  `rstan`
2.  `brms`
3.  `data.table`
4.  `ggplot2`

First, we will load these required packages and configure some useful options.

```{r options, include=T, results='hide', message=F, warning=F}
library(rstan)
library(brms)
library(data.table)
library(ggplot2)

set.seed(10)
options(mc.cores = parallel::detectCores())
theme_set(theme_bw())
```

Now, we will load the functions necessary to fit a custom family in `brms`.

```{r stan_funs, include=T}
source('R/stan_funs.R')
stanvars <- stanvar(scode = stan_funs, block = "functions")
```

This `.R` file loads a `brms` family, two functions, and a string containing Stan functions into the global environment. The custom family, `sinhasinh`, will allow us to fit a model in `brms` with a likelihood that is not included in `brms` by default. It is defined as

```{r sinhFamily, eval=F}
sinhasinh <- custom_family(
  name = "sinhasinh",
  dpars = c("mu", "sigma", "eps", "delta"),
  links = c("identity", "log", "identity", "log"),
  lb = c(NA, NA, NA, NA),
  type = "real"
)
```

In the `dpars` argument, we are defining the names of the parameters of custom family. Note that every custom family in `brms` must have "mu" as its first distributional parameter. With `links`, we are defining the link functions for the linear models for each distributional parameter: both $\sigma$ and $\delta$ must be greater than zero, so we assign them `log` link functions. The `lb` argument sets the lower bounds of each parameter, and the `type` argument defines the support of the distribution.

The R script we ran also adds two functions to the global environment:

1.  `log_lik_sinhasinh`: allows `brms` to calculate the log-likelihood of the sinh-arcsinh distribution using Stan functions that we will expose later on.
2.  `predict_sinhasinh`: allows `brms` to sample from the sinh-arcsinh distribution similarly using exposed functions from Stan.

Both of these functions are named in accordance with `brms` convention (`log_lik_FAMILYNAME` and `predict_FAMILYNAME`) so that we will be able to use the posterior prediction and checking tools built-in to the package.

Finally, the script sets a string called `stan_funs` in the global environment. This string contains two Stan functions, `sinhasinh_lpdf` and `sinhasinh_rng`, which allows to calculate the sinh-arcsinh log-density and take sinh-arcsinh samples, respectively. The log-density function with that particular name is required for the custom family to work. We will expose both functions to R after fitting the first model, and the two R functions defined above will wrap them.

For convenience, we also define a native R function to produce sinh-arcsinh samples.

```{r sinhSampler}
rsinhasinh <- function(n, mu = 0, sigma = 1, eps = 0, delta = 1) {
  mu + sigma * sinh((asinh(rnorm(n, 0, 1)) - eps) / delta)
}
```

## Simulated Data

We will generate sinh-arcsinh distributed data with moments that vary with data to fit to. First, we define the population of interest.

```{r definePop, results='asis'}
N <- 300
ages <- round(runif(N, 15, 64))
pct_f <- 0.6
sexes <- c('Male', 'Female')[rbinom(N, 1, pct_f) + 1]

# Load everything into a dataframe
age_df <- data.frame(age = ages, sex = sexes, id = 1:N)
age_df$scaled_age <- (ages - mean(ages)) / sd(ages)
age_df$sex <- relevel(factor(age_df$sex), ref = 'Female')
age_df$age_bin <- with(age_df, age - age %% 5)
```

Now we will use distribution regression coefficients from Wolock (2021) to predict distributional parameters for each individual. These coefficients come from a model with linear age-sex interactions for all four parameters, so we need to be sure to define the model matrix in the same way.

```{r predictParams}
# Hard-coded coefficients from paper
B_mu <- c(0.1, -0.02, -0.18, 0.01)
B_sigma <- c(-2.44, -0.11, 0.00, 0.26)
B_eps <- c(-0.2, 0.01, 0.36, 0.06)
B_delta <- c(-0.46, -0.05, 0.01, 0.04)

# Create model matrix for simulating data
X <- model.matrix(id ~ scaled_age * sex, data = age_df)

# Get true distributional parameter values
age_df$mu <- X %*% B_mu
age_df$delta <- exp(X %*% B_delta)
age_df$sigma <- exp(X %*% B_sigma) * age_df$delta
age_df$eps <- X %*% B_eps
```

With all four distributional parameters in hand for every individual, we can sample sinh-arcsinh distributed outcomes. In this case, the dependent variable of the model those coefficients are associated with the log-ratio of partner's age to respondent's age.

```{r simOutcome}
# Sample with function we defined earlier
age_df$log_ratio <- with(age_df, rsinhasinh(N, mu, sigma, eps, delta))
# Calculate partner age
age_df$p_age <- round(exp(age_df$log_ratio) * age_df$age)
```

We can plot the resulting distribution for several five-year age bins by sex.

```{r simHist, echo=F, fig.height=4}
ggplot(age_df[age_df$age_bin %in% c(20, 25, 30),],
       aes(x = p_age)) +
  geom_bar() +
  facet_grid(sex ~ age_bin, scales='free_y') +
  coord_cartesian(expand = 0) +
  labs(x = 'Partner age', y = 'Number of observations')
```

## Fitting Models

Now, we will fit three models in `brms`:

1.  **Conventional:** regression with linear age-sex interaction for the location parameter and constant higher-order parameters
2.  **Distributional 1:** distributional regression with linear age-sex interaction for location and independent age and sex effects for higher-order parameters
3.  **Distributional 2:** distributional regression with linear age-sex interactions for all four parameters

### Conventional Regression

First, we fit the conventional model. We use the `bf` function to build a `brms` formula with models for all four distributional parameters in the `sinhasinh` family. Note that we use the first formula (corresponding to `mu`) to set the outcome variable. We include the `stanvars` object defined by `R/stan_funs.R` in the model using the `stanvars` argument of `brm`. (*This cell will take substantially longer to run than the previous ones because the `brm` function compiles a Stan model and performs sampling with HMC.)*

```{r convBRM, results='hide'}
bf_1 <- bf(
  log_ratio ~ scaled_age * sex,
  sigma ~ 1,
  eps ~ 1,
  delta ~ 1
)

brm_1 <- brm(formula = bf_1,
             data = age_df,
             family = sinhasinh,
             stanvars = stanvars)
```

```{r brm1PostSummary}
kable(posterior_summary(brm_1), digits = 2)
```

## References
